Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies

### 2022-09-07 paper.qmd
draft review

---
title: "Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies"
author: "Dean Marchiori"
format: pdf
abstract: "A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data in reported cases. This research aims to collate and standardise available case reports to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data. The analysis identified the classic phenotype for this condition, as well as two other distinct feature clusters. The first relates to findings of vocal cord paralysis and swallowing dysfunction, the other relates to more complex brain anomalies. Furthermore, the survival probability for those affected was found to decline sharply in early infancy, but exhibited a wide range of outcomes provisionally associated with variant type. However, interpretation of these results are guarded based on very low sample sizes. Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
editor: visual
bibliography: refs.bib
execute: 
  echo: false 
---

@@

```{r}
# table with med gen 
# Better tables with gtsummary
```

@@

Neurodevelopment disorders are a diverse and heterogeneous group of conditions impacting the development of the nervous system, brain function, physical development, emotional development and learning ability. A recently identified condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) (MIM:[618622](https://omim.org/entry/618622)) has been described in children with bi-allelic loss-of-function variants in *SMPD4* [@magini2019loss]. Sphingomyelinases such as *SMPD4* play an important celluar role by hydrolyzing sphingomyelin into ceramide and phosphorylcholine. *SMPD4* specifically encodes one of 4 neutral sphinogmyelinases, nSMase3 (MIM: [610457](http://omim.org/entry/610457)). So far, less than 50 cases of this rare disorder have been reported in the literature. Clinical phenotype data of these cases is largely heterogeneous with severe neurological complications and early demise a key feature. This makes the identification, diagnosis and ongoing management of cases a challenge for patients, their families and medical practitioners.

@@

A detailed study by @magini2019loss involved 12 unrelated familes with 32 individuals (21 with detailed clinical information). Presentations were of microcephaly, simplified gyral pattern of the cortex, hypomyelination, cerebellar hypoplasia, congenital arthrogryposis, and early fetal/postnatal demise. Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly. Three missense changes were noted in the study with affected children in these families often showing a milder presentation suggestive of possible residual function. In these cases individuals were able to develop independent motor skills, have mild intellectual disability and arthrogryposis without evidence of simplified gyral patterns on brain MRI. Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant clinical features were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.

@@

To overcome these data challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell. In this case detailed text-based data can be tranformed to high dimensional binary indicators of key clinical features. This structure is conducive to effective statistical analysis and integrates deliberately with the *tidyverse* [@tidyverse] collection of packages for data analysis in R [@rbase].

@@

An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in SMPD4. With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood. Furthermore, the survival time of affected children is not well understood despite early-demise featuring as a severe outcome in many cases. While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.

@@

This will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better understand this newly described condition. This research will assist clinicians understand the typical and non-typical presentations of this challenging new condition. In addition, genetic counselling of families with affected children will benefit from enhanced analysis on outcomes of existing reported cases.

@@

The data were read into R [@rbase] without any changes, so as to preserve the reproducibility of the data transformation steps.

The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10]. This requires each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item).

This resulted in a long-formatted data set of 21 rows (individuals) and 64 columns (clinical features).

In many cases, key clinical information was entered as free-text descriptions, which rendered any attempt of meaningful analysis impractical (@tbl-nontidy). In these cases, the text was tokenised by separating the list of clinical observations at each comma and forming a binary indicator column noting its presence '1' or absence '0' (@tbl-tidy).

|                     | Family 1- Individual 1                                                                                     | Family 1- Individual 4                                      |
|------------------|--------------------------|----------------------------|
| Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |

@@

| id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
|-----------------|-----------|-----------|-----------|-----------|-----------|
| Family 1- Individual 1 | 1                        | 1          | 1              | 0                 | ... |
| Family 1- Individual 4 | 1                        | 0          | 0              | 1                 | ... |

@@

In the case where two variables were formed from phenotypes that are considered to be synonymous, these were merged into one indicator column to prevent duplication e.g. {bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b\_l} -\> {bilateral_cleft_lip}.

@@

The final dataset consisted of 28 observations (one per individual) and 152 variables (one per clinical feature). These variables were comprised of qualitative text descriptions, binary indicators of features as described above and other fields including background information on the patient and various biomarkers. A full list of this data set is available in @tbl-dict. These data sets were compiled into a publicly available R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing.

@@

We let these data be $\mathbf{X}$ which is comprised of a set of individuals $I$ and a set of features $Q$ such that the $q_{th}$ feature has $K_q$ levels. The sum of all categories $K = \sum_{q=1}^{Q}K_q$ defines the dimensionality of $\mathbf{X}$ as an $I\times K$ matrix. This resulted in a dataset $X_{28 \times 61}$ indicator matrix of 61 clinical features across 28 individuals. Taking $\delta_{ik} = 1$ if subject $i$ has feature $k$ and $\delta_{ik} = 0$ if the subject does not, we are left with the completely disjnuctive table $\mathbf{X} = I\times K$ of $\{0,1\}$. Letting the sum of all entries of $\mathbf{X}$ be $N$, we have $\mathbf{Z} = N^{-1}\mathbf{X}$. We can introduce two diagonal matricies $\mathbf{D_r} = diag(\mathbf{r})$ and $\mathbf{D_c} = diag(\mathbf{c})$ where $\mathbf{r}$ and $\mathbf{c}$ are the vectors of row sums and column sums of $\mathbf{Z}$ respectively.

@@

The *contribution* of each category is computed for each of the principal axis per @le2010multiple. This metric identifies the proportion of variance of that axis due to the point and is defined as:

@@

The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features that were internally homogeneous and externally heterogeneous. This was computed in R using the clValid package [@clvalid].

A range of clustering algorithms were comparied including AGNES hierarchical agglomerative [@kaufman2009finding], k-means [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self]. A sensible range of clusters from 2-5 were trialed for each clustering algorithm. The results for each method were evaluated using three measures of stability: Average proportion of non-overlap (APN), Average Distance (AD) and Average Distance Between Means (ADM) [@datta2003comparisons]. Stability measures compare the full clustering result with a result based on dropping meach of the columns one at a time. These metrics were selected as we are representing each data point (clinical feature) with a number of MCA dimensions. A good cluster solution is one that is robust and meaningful across all of these dimensions [@fig-clvalid].

The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics and a sensible partioning of the data for this analysis [@tbl-eig].

@@

To model survival probability for all subjects in the combined data, the data are subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric indicator which is equal to 1 if the subject is deceased and 0 otherwise. The Kaplan-Meier estimator [@kaplan1958nonparametric] is used to estimate the survival function of the data.

@@

$$
\hat{S}(t) = \prod_{t_i \leq t}\frac{n_i - d_i}{n_i} 
$$ where $t_i$ is some event time, $d_i$ represents the number of events (here deceased subjects) and $n_i$ indicating the indivudals known to have survived up to $t_i$

The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book]. Next the Kaplan-Meier estimator was calculated, stratified by the reported type of genetic variation reported in the literature. A logrank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* R package.

Finally, a Cox proportional hazards regression model [@andersen1982cox] was fit on survival time with the variant type as the sole covariate. This was to detect and measure the Hazard ratio of variant type on the survival time in order to test the hypothesis that misense variant types exhibit significantly greater survival compared to other variant types.

@@

The initial results of the MCA were inspected with a bi-plot and identified two subjects (17, 18) from the data that were clear outliers and sat significatly beyound the 95% confidence ellipse. These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and were subject to a missense variation in *SMPD4*. Further analysis of this influence is conducted below. For now, these subjects were removed to allow for more meaningful analysis.

@@

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplotprelim
#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables. A 95% confidence ellipse is plotted in red around the origin."
fviz_mca_biplot(
  mca_prelim,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point", "text"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.2, contrib = NULL),
  palette = "grey",
  col.ind = "red",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

@@

The *contribution* of the categories were ranked within each of the first four MCA dimensions See @fig-ctr. The first dimension is dominated by highly specific facial dysmophisms such as epicanthal folds, short philtrum and low anteriour hairline.

@@

Examining the bi-plot of the first two principal axes @fig-biplot shows contrasting sources of variation between conditions relating to airway and swallowing function and predominantly structural brain anomalies such as as cerebellar hypoplasia, agenesis of corpus collosum and development delay.

The third dimension is characterised by complex features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features. These are contrasted by the fourth principal dimension which is typical for respiratory distress including stridor and hypercapnia.

```{r ctr-plot, fig.height=5, fig.width = 8, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 10 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(10, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered()+
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  theme(legend.position = 'bottom') +
  labs(y = '')
```

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplot
#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplot2
#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey cicles indicating the individual's projected position and black triangles indicaating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(3, 4),
  geom = c("point", "text"),
  geom.ind = "point",
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)

@@

Internal cluster validation measures resulted in between 2 and 6 clusters as the optimal choice.

Proceeding with the optimal cluster selection, per the Silhouette coefficient of 6 clusters.

A scatterplot of the cluster results @fig-clust projected on just the first two dimensions highlights a core cluster of classic features of NEDMABA. A separate cluster is formed of airway and feeding related conditions involing vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.

Some smaller satellite clusters exists where highly distinctive features were associated with an indivual or family. Also a singleton clusters exist for features that are distinct or unusual such as posteriorly angulated ears.

A full articulation of feature to cluster assignment is given in @tbl-profile.

@@

```{r, warning=FALSE}
#| label: fig-clvalid
#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters. The most optimal number of clusters are selected by picking the lowest value from one or many of these compariable metrics."

@@

```{r, fig.height=6, fig.width=6,  warning=FALSE}
#| label: fig-clust
#| fig-cap: "A projection of features on the first two principal components with k-means cluster assignment indicated by colour."

d <- cbind(cluster_data, clust = factor(cluster_assignments)) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  arrange(var, clust)

@@

  ggplot(d, aes(dim_1, dim_2, colour = clust)) +
  geom_polygon(data = hull, alpha = 0.2, aes(fill = clust)) +
  geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
  geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=4), aes(label = var), size = 3, show.legend = F) +
  geom_point(aes(colour = clust), position = "jitter", show.legend = F) +
  labs(x = "Dimension 1",
       y = "Dimension 2",
       fill = 'cluster') +
    guides(color='none')
```

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with survival probability at time 0 of 0.82 (0.69, 0.98). Survival probability after one year at 0.403 (0.252, 0.644) with median survival of 150 days and survival probability at close to ten years of 0.24 (0.11, 0.52).

```{r}
# KM Curve
x <- survminer::surv_summary(kapm_baseline, data = survival_data)
kable(x, digits = 2, caption = "Survival probabilities derived from Kaplan-Meier estimation. ")
```

```{r, warning=FALSE}
#| label: fig-surv
#| fig-cap: "Kaplan Meier curve with the black line representing the survival probability and with grey shading indicating the 95% confidence interval."
survminer::ggsurvplot_df(x,
                         color = "black",
                         conf.int = TRUE,
                         conf.int.fill = "gray",
                         risk.table = TRUE, # Add risk table
                         risk.table.col = "strata", # Change risk table color by groups
                         linetype = "strata", # Change line type by groups
                         surv.median.line = "hv", # Specify median survival
                         ggtheme = theme_bw(),
                         legend = "none") +
  labs(x = "time (days)")

@@

A comparitive boxplot of survival times by variant type shows significant outliers for missense+splice and missense variations. However, even within missense hymozygous variants the survival time varies greatly. Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as noted by @magini2019loss.

>>>>

---
title: "Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies"
author: "Dean Marchiori"
format: pdf
abstract: "A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data in reported cases. This research aims to collate and standardise available case reports to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data. The analysis identified the classic phenotype for this condition, as well as one other distinct feature cluster relating to findings of vocal cord paralysis and respiratory failure. The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some as high as 12.5 years. This wide range of outcomes is provisionally associated with variant type however interpretation of these results are guarded based on very low sample sizes. Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
editor: visual
bibliography: refs.bib
execute: 
  echo: false 
---

@@

```{r}
# table with med gen 
```

@@

Neurodevelopment disorders are a diverse and heterogeneous group of conditions impacting the development of the nervous system, brain function, physical development and learning ability. A recently identified condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) (MIM:[618622](https://omim.org/entry/618622)) has been described in children with bi-allelic loss-of-function variants in *SMPD4* [@magini2019loss]. Sphingomyelinases such as *SMPD4* play an important cellular role by hydrolyzing sphingomyelin into ceramide and phosphorylcholine. *SMPD4* specifically encodes one of 4 neutral sphinogmyelinases, nSMase3 (MIM: [610457](http://omim.org/entry/610457)). So far, less than 50 cases of this rare disorder have been reported in the literature. Clinical phenotype data of these cases is largely heterogeneous with severe neurological complications and early demise a key feature. This makes the identification, diagnosis and ongoing management of cases a challenge for patients, their families and medical practitioners.

@@

A detailed study by @magini2019loss involved 12 unrelated families with 32 individuals (21 with detailed clinical information). Presentations were of microcephaly, simplified gyral pattern of the cortex, hypomyelination, cerebellar hypoplasia, congenital arthrogryposis, and early fetal/postnatal demise. Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly. Three missense changes were noted in the study with affected children in these families often showing a milder presentation suggestive of possible residual function. In these cases individuals were able to develop independent motor skills, have mild intellectual disability and arthrogryposis without evidence of simplified gyral patterns on brain MRI. Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant clinical features were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.

@@

To overcome these data challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell. In this case detailed text-based data can be transformed to high dimensional binary indicators of key clinical features. This structure is conducive to effective statistical analysis and integrates deliberately with the *tidyverse* [@tidyverse] collection of packages for data analysis in R [@rbase].

@@

An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in *SMPD4*. With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood. Furthermore, the survival time of affected children has not been extensively analysed despite early-demise featuring as a severe outcome in many cases. While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.

@@

This will be the first in-depth analysis of this novel variant, which aims to produce statistical findings to better understand this newly described condition. This research will assist clinicians understand the various presentations of this challenging new condition. In addition, genetic counselling of families with affected children will benefit from enhanced analysis on outcomes of existing reported cases.

@@

The data were read into `R` [@rbase] without any changes, so as to preserve the reproducibility of the data transformation steps. The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10]. This requires each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item). This resulted in a long-formatted data set of 21 rows (individuals) and 64 columns (clinical features). In many cases, key clinical information was entered as free-text descriptions, which rendered any attempt of meaningful analysis impractical (@tbl-nontidy). In these cases, the text was tokenised by separating the list of clinical observations at each comma and forming a binary indicator column noting its presence '1' or absence '0' (@tbl-tidy).

|                     | Family 1- Individual 1                                                                                     | Family 1- Individual 4                                      |
|---------------|-------------------------------------|--------------------|
| Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |

@@

| id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
|------------|-------------|------------|------------|------------|------------|
| Family 1- Individual 1 | 1                        | 1          | 1              | 0                 | ... |
| Family 1- Individual 4 | 1                        | 0          | 0              | 1                 | ... |

@@

In the case where two variables were formed from features that are considered to be synonymous, these were merged into one indicator column to prevent duplication e.g. {bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b\_l} -\> {bilateral_cleft_lip}.

@@

The final data set consisted of 28 observations (one per individual) and 152 variables (one per clinical feature). These variables were comprised of qualitative text descriptions, binary indicators of features as described above and other fields including background information on the patient and various bio markers. A full list of this data set is available in @tbl-dict. These data sets were compiled into a publicly available `R` Package called `SMPD4` [@smpd4-data] in order to allow for reproducibility and sharing.

@@

We let these data be $\mathbf{X}$ which is comprised of a set of individuals $I$ and a set of features $Q$ such that the $q_{th}$ feature has $K_q$ levels. The sum of all categories $K = \sum_{q=1}^{Q}K_q$ defines the dimensionality of $\mathbf{X}$ as an $I\times K$ matrix. This resulted in a data set $X_{28 \times 61}$ indicator matrix of 61 clinical features across 28 individuals. Taking $\delta_{ik} = 1$ if subject $i$ has feature $k$ and $\delta_{ik} = 0$ if the subject does not, we are left with the completely disjunctive table $\mathbf{X} = I\times K$ of $\{0,1\}$. Letting the sum of all entries of $\mathbf{X}$ be $N$, we have $\mathbf{Z} = N^{-1}\mathbf{X}$. We can introduce two diagonal matrices $\mathbf{D_r} = diag(\mathbf{r})$ and $\mathbf{D_c} = diag(\mathbf{c})$ where $\mathbf{r}$ and $\mathbf{c}$ are the vectors of row sums and column sums of $\mathbf{Z}$ respectively.

@@

The *contribution* of each category was computed for each of the principal axis per @le2010multiple. This metric identifies the proportion of variance of that axis due to the point and is defined as:

@@

The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features that were internally homogeneous and externally heterogeneous. This was computed in R using the `clValid` package [@clvalid].

A range of clustering algorithms were compared including AGNES hierarchical agglomerative [@kaufman2009finding], k-means [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self]. A sensible range of clusters from 2-5 were trialed for each clustering algorithm. The results for each method were evaluated using three measures of stability: Average proportion of non-overlap (APN), Average Distance (AD) and Average Distance Between Means (ADM) [@datta2003comparisons]. Stability measures compare the full clustering result with a result based on dropping each of the columns one at a time. These metrics were selected to represent each data point (clinical feature) with a number of MCA dimensions. A good cluster solution is one that is robust and meaningful across all of these dimensions [@fig-clvalid]. The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics and a sensible partitioning of the data for this analysis [@tbl-eig].

@@

To model survival probability for all subjects in the combined data, the data were subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric event indicator which is equal to 1 if the subject is deceased and 0 if the subject was still alive at the time the respective study had concluded.  The Kaplan-Meier estimator [@kaplan1958nonparametric] was used to estimate the survival function of the data.

@@

$$
\hat{S}(t) = \prod_{t_i \leq t}\frac{n_i - d_i}{n_i} 
$$ where $t_i$ is some event time, $d_i$ represents the number of events (here deceased subjects) and $n_i$ indicating the individuals known to have survived up to $t_i$

The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book]. Next the Kaplan-Meier estimator was stratified by the type of genetic variation reported in the literature. A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* R package.

@@

The initial results of the MCA were inspected with a bi-plot and identified two subjects (17, 18) from the data that were clear outliers and sat significantly beyond the 95% confidence ellipse. These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and were subject to a missense variation in *SMPD4*. Further analysis of this influence is conducted below. For now, these subjects were removed to allow for more meaningful analysis.

@@

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplotprelim
#| fig-cap: "A bi-plot of the first two principal axes with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in red around the origin."
fviz_mca_biplot(
  mca_prelim,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point", "text"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.2, contrib = NULL),
  palette = "grey",
  col.ind = "red",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

@@

The *contribution* of the categories were ranked within each of the first four MCA dimensions (@fig-ctr). The first dimension is dominated by highly specific facial dysmorphisms such as epicanthal folds, short philtrum and low anterior hairline.

@@

Examining the bi-plot of the first two principal axes @fig-biplot illustrates the contrasting sources of variation between conditions relating to airway and swallowing function and predominantly facial features and structural brain anomalies such as as cerebellar hypoplasia, development delay.

The third dimension is characterised by atypical, complex features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features. These are contrasted by the fourth principal dimension which is strongly categorised by respiratory distress including stridor and hypercapnia in the absence of typical presentations of simplified gyral pattern, intrauterine growth retardation and microcephaly.

```{r ctr-plot, fig.height=6, fig.width = 8, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 10 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(10, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  mutate(vars = str_remove(vars, "_[:digit:]")) |> 
  mutate(vars = str_remove(vars, "_ind")) |> 
  mutate(vars = str_replace_all(vars, "_", " ")) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered()+
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  theme(legend.position = 'bottom') +
  labs(y = '')
```

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplot
#| fig-cap: "A bi-plot of the first two principal axes with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplot2
#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey circles indicating the individual's projected position and black triangles indicating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(3, 4),
  geom = c("point", "text"),
  geom.ind = "point",
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)

@@

Cluster validation measures resulted in either 2 or 5 clusters as the optimal choice.

In order to partition the data into meaningful groupings, k-means with 5 clusters was selected. This reflects the optimal choice for both AD and FOM cluster metrics.

A scatter plot of the cluster results @fig-clust projected on just the first two dimensions highlights a core cluster of classic features of NEDMABA such as microcephaly, hypomyelination, arthrogryposis and feet deformity and structural brain anomalies. A separate cluster is formed of airway and feeding related conditions involving vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.

Three other smaller satellite clusters exists where highly distinctive features were associated with an individual or family. A full articulation of feature to cluster assignment is given in @tbl-profile.

@@

```{r, warning=FALSE}
#| label: fig-clvalid
#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters. The most optimal number of clusters are selected by picking the lowest value from one or many of these comparable metrics."

@@

```{r, fig.height=6, fig.width=6,  warning=FALSE}
#| label: fig-clust
#| fig-cap: "A projection of features on the first two principal components with k-means cluster assignment indicated by colour. A random sample of three labels per cluster are printed to provide a representation of each cluster's characteristics."

d <- cbind(cluster_data, clust = factor(cluster_assignments)) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  arrange(var, clust) |> 
  mutate(clust = fct_recode(clust,
    Core = "4",
    Adjacent = "2",
    Outlying_3 = "3",
    Outlying_5 = "5",
    Outlying_1 = "1",
    ))

@@

  ggplot(d, aes(dim_1, dim_2, colour = clust)) +
  geom_polygon(data = hull, alpha = 0.2, aes(fill = clust)) +
  geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
  geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=3), aes(label = var), size = 3, show.legend = F) +
  geom_point(aes(colour = clust), position = "jitter", show.legend = F) +
  labs(x = "Dimension 1",
       y = "Dimension 2",
       fill = 'cluster') +
    guides(color='none')
```

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days. At time 0 the survival probability was 0.82 (0.69, 0.98). Survival probability after one year at 0.403 (0.252, 0.644) and survival probability at close to ten years of 0.24 (0.11, 0.52).

```{r}
# KM Curve
x <- survminer::surv_summary(kapm_baseline, data = survival_data)
kable(x, digits = 2, caption = "Survival probabilities derived from Kaplan-Meier estimation for the range of recorded survival times, including 95% Confidence Interval.")
```

```{r, warning=FALSE}
#| label: fig-surv
#| fig-cap: "Kaplan Meier curve with the black line representing the survival probability and with grey shading indicating the 95% confidence interval. Cross icons represent right-censoring"
survminer::ggsurvplot_df(x,
                         color = "black",
                         conf.int = TRUE,
                         conf.int.fill = "gray",
                         risk.table = TRUE, # Add risk table
                         risk.table.col = "strata", # Change risk table color by groups
                         linetype = "strata", # Change line type by groups
                         surv.median.line = "hv", # Specify median survival
                         ggtheme = theme_bw(),
                         legend = "none") +
  labs(x = "time (days)")

@@

A comparative box plot of survival times by variant type shows significant outliers for missense+splice and missense variations. However, even within missense homozygous variants the survival time varies greatly. Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as noted by @magini2019loss.
